Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-06-14 Manuscript/manuscriptPS.tex
Update abstract after discussion in issue https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/94

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain poorly understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
We then use this data set
to assess,
in a wide range of available molecular dynamics (MD) simulation models (force fields) of PS,
the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes.
We find that none of the MD models reproduce the NMR data within experimental accuracy.
The best MD models, however, suggest that the carboxyl group in the PS headgroup
does not rotate freely. In line with our previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity. Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results pave the way for the improvement of lipid force fields that correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

>>>>

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain poorly understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
We then use this data set
to assess the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes
in a wide range of available molecular dynamics (MD) force fields of PS.
We find that none of the MD simulations reproduce the NMR data within experimental accuracy.
However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
In line with our previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity. Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results pave the way for the improvement of lipid force fields that correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}
